---
title: "From sequences to trees with Bd Fluidigm data"
output: github_document
editor_options: 
  chunk_output_type: console
---

For this workshop I will show you how to take our Bd Fluidigm data and turn it into a phylogenetic tree. Fun!

First, some preparation

1. I've included the program [Astral](https://github.com/smirarab/ASTRAL) in the workshop github folder. Navigate to the workshop folder and unzip the .zip file.

2. Download the program newick utils. Go to the workshop folder and unzip it. Then navigate to the folder that has the newick utils in terminal and run the following to make the binaries we will need to run:

> ./configure
> make

Ok now let's make sure we have the libraries we need and load them:

```{r setup, include=FALSE}

if(!require(knitr)){
    install.packages("knitr")
    library(knitr)
}
if(!require(BiocManager)){
    install.packages("BiocManager")
    library(BiocManager)
}
if(!require(Biostrings)){
    BiocManager::install("Biostrings")
    library(Biostrings)
}
if(!require(tidyverse)){
    install.packages("tidyverse")
    library(tidyverse)
}
if(!require(muscle)){
    BiocManager::install("muscle")
    library(muscle)
}
if(!require(filesstrings)){
    install.packages("filesstrings")
    library(filesstrings)
}
if(!require(AstralPlane)){
    devtools::install_github("chutter/AstralPlane")
    library(AstralPlane)
}
if(!require(ape)){
    install.packages("ape")
    library(ape)
}

```

Now let's read in the sequence data contained in the "seq_data" files. Each fasta file represents an individual Bd sample. This folder includes Bd samples collected from around Pymatuning, in addition to some reference Bd samples to contextualize our resutls. These fasta files have one amplicon sequence per locus that has multiple alleles coded using [IUPAC ambiguity codes](https://droog.gs.washington.edu/parc/images/iupac.html). 


```{r}
#get a vector of all the file names in the folder
files_PA <- list.files("./seq_data", full.names = T)

```

Now let's see how many sequences we have for each primer/sample.

```{r}
#make empty objects
length_table_PA <- tibble(sample = "", length=0, file="")

#extract sample names
samplenames_PA <- sapply(files_PA, function(x)unlist(strsplit(x,split=".fasta_primerfix.fasta_trim"))[1])
samplenames_PA <- sapply(samplenames_PA, function(x)unlist(strsplit(x,split="./seq_data/Sample."))[2])

#count number of sequences per sample
for (i in 1: length(files_PA)){
    seq <- readDNAStringSet(files_PA[i], format = "fasta")
    length_table_PA <- add_row(length_table_PA, sample=samplenames_PA[i], length=length(seq), file=files_PA[i])
}

#plot the results
hist(as.numeric(length_table_PA$length[-1]), breaks=50,main="PA", xlab="Number of sequences")

```

We can see that each sample has a varible amount of missing data. At this stage in the process we would choose what cutoff we would like to use for including samples. 

```{r}
#get list of only samples with at least 100 sequences 
files_PA_90 <- filter(length_table_PA, length>100)

filenames_PA_90 <- files_PA_90$file
length(filenames_PA_90)

#filters out 1 bad sample
```

Ok now that we have chosen our samples, let's get our data organized. We are going to do that by creating a matrix m that we will then populate with our sequence data. Here each row is a sample and each column is a single locus

```{r}

#make empty matrix of the right size
m_PA <-matrix(NA, nrow=nrow(files_PA_90), ncol=243)
#rename the columns for the loci
colnames(m_PA) <- seq(1,243)
#rename the rows for the samples
rownames(m_PA) <- files_PA_90$sample
#covert
seq_matrix <- as_tibble(m_PA)

#populate the empty matrix with the ambiguities sequences for PA
for (i in 1:length(filenames_PA_90)){
  seq <- readDNAStringSet(filenames_PA_90[i])
  for (j in 1:length(seq)){
    col_match <- as.numeric(names(seq))
    m_PA[i,col_match[j]] <- as.character(seq[j])
     }
}

#here might be a cool place to view the m_PA matrix to get an idea of the data we are working with.

#in Rstudio
#View(m_PA)

```

Now we have a matrix *m_PA* with samples as rows and primers as columns. First we can eliminate primers with no data. Let's get the average length of each sequence to determine which has no data. We can also find the min max and mean sequence length to identify potential bad sequences. 


```{r, include=FALSE}
#create an empty object
n_bases_PA <- matrix(NA, nrow=ncol(m_PA), ncol=7)

#characterize each locus
for (i in 1:ncol(m_PA)){
  n_bases_PA[i,2] <- mean(nchar(m_PA[,i]),na.rm=T)
  n_bases_PA[i,3] <- min(nchar(m_PA[,i]),na.rm=T)
  n_bases_PA[i,4] <- max(nchar(m_PA[,i]),na.rm=T)
  n_bases_PA[i,5] <- median(nchar(m_PA[,i]),na.rm=T)
  n_bases_PA[i,6] <- max(nchar(m_PA[,i]),na.rm=T) - min(nchar(m_PA[,i]),na.rm=T)
  n_bases_PA[i,7] <- sum(is.na(m_PA[,i]))
}
#we get some warnings here but nothing to worry about, it is just becasue some loci are totally missing

#rename columns and add amp labels
n_bases_PA[,1] <- colnames(m_PA)
colnames(n_bases_PA) <- c("amp","mean","min","max","median","diff","sum_missing")

```

Now let's explore the problem amps/loci. These are ones missing too much data.

```{r}
#identify amps with more than 1/2 missing data - only considering the samples of interest and not the reference samples (hence the +13)
badamps_PA <- which(as.numeric(n_bases_PA[,7]) > (.5*(length(filenames_PA_90)+13)))

#now trim out these amos fro the matrix
m_trim_PA <- m_PA[,-badamps_PA]

#check how many we are left with
dim(m_trim_PA) 

```

For each sample with data, aligns all loci separately. This is done only for the gene tree to species tree approach.

```{r, include=FALSE}
#run once
dir.create("PA_locus_align")

#aligns all seqs (removing those with NAs) and writes them out as a separate fasta into their respective directories 

for (i in 1:ncol(m_trim_PA)){
  locus_list <- m_trim_PA[!is.na(m_trim_PA[,i]),i]
  locus_seq <- DNAStringSet(locus_list)
  locus_align <- muscle::muscle(locus_seq)
  locus_align <- DNAStringSet(locus_align)
  writeXStringSet(locus_align, paste(colnames(m_trim_PA)[i], "_align.fasta", sep=""))
  file.move(paste(colnames(m_trim_PA)[i], "_align.fasta", sep=""), "./PA_locus_align")
}

```

Now we want to make a tree for each locus. This allows us to collapse all the information contained in one locus into the relative relationships between the samples. These trees will then be used to create another tree using the program ASTRAL-III. To do this I typically use the program geneious (as I will demonstrate) to run the command line program raxml to make a maximum liklihood tree with bootstrap support. Raxml can also be run locally after installing it. This installation can be a bit tricky and making the trees can be time consuming so I'll show you how I do it then provide the trees. Really you can use any program that makes trees with bootstrap support. 

Now for this project I loaded the alignments for each locus into the program geneious. From there I ran raxml for each locus using the following command: 

> raxmlHPC-SSE3-MAC -s input.phy -n output -m GTRCAT -f a -x 1 -N 100 -p 1

You can also run raxml [online](https://raxml-ng.vital-it.ch/#/). Lots of options!



Next I concatenated all newick trees into one file to be input into Astral. In the directory with all the trees (and only the trees) with nodes collapsed I ran the following:

for mac:
> cat * > combined_trees.newick

for windows:
> copy /b *.newick combined_trees.newick

Here is code that will run this command right from R.

```{r}
system("cat PA_locus_trees/* > PA_trees_combined.newick")
```

Then we can use the newick utils tools to collapse all nodes with <10 bootstrap support. Make sure you have extracted the newick utils zipped file and run the compile and make commands (see above).

```{r}
system("./newick-utils-1.6/src/nw_ed PA_trees_combined.newick 'i & b<=10' o > PA_trees_collapse10.newick")
```

Then run astral on your set of collapsed trees. Double check that you have extracted the zipped Astral file and that the path is correct. 

```{r}
system("java -jar ./Astral/astral.5.7.5.jar -t 2 -i PA_trees_collapse10.newick -o PA_astral2.tre")
```

Now we have a TREE!! Let's look at it! Here we can use a handy program called [AstralPlane](https://github.com/chutter/AstralPlane) to display our tree

```{r}

astral.data = createAstralPlane(astral.tree = "PA_astral2.tre",
                                outgroups = "Bd_RFM380",
                                tip.length = 1)


astralProjection(astral.plane = astral.data,
                 local.posterior = TRUE,
                 pie.plot = TRUE,
                 pie.data = "genetree",
                 save.file = "astral_plot.pdf",
                 pie.colors = c("purple", "blue", "green"),
                 node.color.text = c("white"),
                 node.color.bg = c("black"),
                 node.label.size = 0.3,
                 tip.label.size = 0.5,
                 pie.chart.size = 0.75)

```

Here we have a lot of low support nodes showing up. Let's go ahead and collapse some of the low support nodes and then re-display our tree with support values. Here we will run Astral one more time to use a different notation (-t 3) use newick utils again to collapse low support nodes.

```{r}

system("java -jar ./Astral/astral.5.7.5.jar -t 3 -i PA_trees_collapse10.newick -o PA_astral3.tre")

#collapse nodes with <0.5 posterior support
system("./newick-utils-1.6/src/nw_ed PA_astral3.tre 'i & b<=0.5' o > PA_astral_collapse5.tre")

```

Read in the new astral tree and plot it.

```{r}

astral_collapse <- read.tree("PA_astral_collapse5.tre")
astral_collapse <- root(astral_collapse, "Bd_RFM380")

plot(astral_collapse, show.node.label=T)

```

Now it is probably helpful to read in a metadata file that has more information about each sample so we can be more specific with our labels.

```{r}
#read in raw metadata file
PA_meta <- read.csv("PA_samples_metadata.csv")

#pull out names from the tree
samples <- as.data.frame(astral_collapse$tip.label)
colnames(samples) <- "Sample_ID"
#now join the info from the metatable to the sample names of the tree
sample_info <- left_join(samples, PA_meta, by = "Sample_ID")

#now make a copy of the tree so we can change the tip labels
astral_collapse_species <- astral_collapse
astral_collapse_species$tip.label <- sample_info$Species_code
refs <- which(astral_collapse_species$tip.label == "")
astral_collapse_species$tip.label[refs] <- sample_info$GENOASSIGN[refs]

#now plot
plot(astral_collapse_species, show.node.label=T, tip.color=as.numeric(as.factor(sample_info$GENOASSIGN)), cex=0.6, main="Astral Tree for PA Bd Samples")

```


